Bayesian Statistical Modeling with Predictors from LLMs

### 2024-04-05 04-paper/paper-LLM-probmod.tex
revised section 4

An LLM is design to predict the next token given an input string.
This is essentially an item-level prediction: next-token probabilities are specified for concrete strings after a concrete instance of a task, i.e., an item of the task, not for the task as such.
From these next-token probabilities, we can derive probabilities for multiple-choice answers for each item (this section) and for a condition (next section).

Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
% For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ tokens, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same response category $r_{i}$.
For the case at hand these categories are: target, competitor, distractor, so that $y_{k1}$ always corresponds to the designated \emph{target option}, $r_{1}$, that is the ``gold standard'' answer.

The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
  \mf{TODO: Add quick word on how this can in principle be extended to masked language models.}
  More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
  From the point of view of experimental psychology, these corrections are \emph{post hoc} fixes to improperly balanced experimental materials.
  For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
  \mf{check whether this is correct also after new LLaMA results.}
}
%
\begin{align*}
  % S\left( y_{ki}, x_{k} \right) =
  \text{S}_{ki}
  % = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right)
  =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
\end{align*}
%

@@

As mentioned previously, the items of the reference game experiment from Section~\ref{experiment-reference-games} differ in which levels of features (color, shape, texture) instantiate the structure of the task shown in Figure \ref{fig:ref-game}, as well as the order of presentation of objects and words.
We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, such as over-production of informationally irrelevant material \citep[e.g.,][]{DaviesKatsos2010:Over-informativ,Rubio-Fernandez2019:Overinformative,DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options. \mf{references prompt variation}
It therefore becomes an empirical question of whether item-level predictions from LLMs provide a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.

To address this question, we assessed item-level scores $\text{S}_{ki}$ from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} for a text-based version of the reference game from Section~\ref{experiment-reference-games}.
% \footnote{
  % Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
The choice options are categorized as in the human experiment.\footnote{
  In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option.}
Similar to the RSA model fit, we allow for random errors with probability $\epsilon_{C}$:
%
\begin{align*}
  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
\end{align*}
%
The resulting likelihood function for item-level data with item-level predictions is, where priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.:
%
\begin{align*}
 P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
\end{align*}
%

Summary statistics for samples from the posterior distribution over parameters are shown in Table~\ref{tab:sumStats} (rows 4 and 5).
Credible values of $\alpha$ parameters are comparatively low for the LLM-based model, suggesting that the predicted scores have rather large differences, which have to be compensated for by the softmax link-function to adequately fit the data.
Figure~\ref{fig:closeness-target-item-level} shows the mean of the probabilities for the highest scoring answer category, averaged over all items, for different parameter values of $\alpha$, highlighting that for \textit{a posteriori} credible values of $\alpha$ (gray shaded areas), the predictions are clearly distinct from the predictions of a WTA strategy (which assigns probability 1 to the highest scoring option for each item).
This suggests that the WTA strategy, which is the special limiting case for $\alpha \rightarrow \infty$, provides a substantially worse fit for item-level choices than those provided by less extreme values of $\alpha$.

@@

@@

Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated LLM scores are inadequate predictors of the human data.
To corroborate this conclusion, Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
Concretely, the plots show, for each condition and item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
So, the model itself predicts systematic variability at the item level.
The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
Put more strongly, a model that uses the most obvious item-level score to predict what human participants choose on a by-item level is ruled out by the current experimental data.

>>>>

An (autoregressive) LLM is designed to predict the next token given an input string.\footnote{
  Nothing of substance changes when a Bayesian statistical model is built around scores from a masked language model.
  We focus on autoregressive, left-to-right language modeling and next-token prediction for ease of reference, since all models tested here are autoregressive.
}
This is essentially an item-level prediction: next-token probabilities are specified for concrete strings after a concrete instance of a task, i.e., an item of the task, not for the task as such.
From these next-token probabilities, we can derive probabilities for multiple-choice answers for each item (this section) and for a condition (next section).

\paragraph{Notation.}
Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
% For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ tokens, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same response category $R_{i}$.
For the case at hand these categories are: target, competitor, distractor, so that $y_{k1}$ always corresponds to the designated \emph{target option}, $R_{1}$.

\paragraph{LLM scores \& predictions.}
The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
  More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
  From the point of view of experimental psychology, these corrections are \emph{post hoc} fixes to improperly balanced experimental materials.
  For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
  \mf{@PT: check whether this is correct also after new LLaMA results.}
}
%
\begin{align*}
  % S\left( y_{ki}, x_{k} \right) =
  \text{S}_{ki}
  % = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right)
  =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
\end{align*}
%

@@

\paragraph{Bayesian statistical modeling with LLM predictors.}
As mentioned previously, the items of the reference game experiment from Section~\ref{experiment-reference-games} differ in which levels of features (color, shape, texture) instantiate the structure of the task shown in Figure \ref{fig:ref-game}, as well as the order of presentation of objects and words.
We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, such as over-production of informationally irrelevant material \citep[e.g.,][]{DaviesKatsos2010:Over-informativ,Rubio-Fernandez2019:Overinformative,DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options. \mf{references prompt variation}
It therefore becomes an empirical question of whether item-level predictions from LLMs provide a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.

To address this question, we assessed item-level scores $\text{S}_{ki}$ from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} for a text-based version of the reference game from Section~\ref{experiment-reference-games}.
% \footnote{
  % Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
The choice options are categorized as in the human experiment.\footnote{
  In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option.}
Similar to the RSA model fit, we allow for random errors with probability $\epsilon_{C}$:
%
\begin{align*}
  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
\end{align*}
%
The resulting likelihood function for item-level data with item-level predictions is, where priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}:
%
\begin{align*}
 P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le i \le 3} \right )\,.
\end{align*}
%

Summary statistics for samples from the posterior distribution over parameters are shown in Table~\ref{tab:sumStats} (rows 4 and 5).
Credible values of $\alpha$ parameters are comparatively low for the LLM-based item-level model, suggesting that the predicted scores have rather large differences, which have to be compensated for by the softmax link-function to adequately fit the data.
Figure~\ref{fig:closeness-target-item-level} shows the mean of the probabilities for the highest scoring answer category, averaged over all items, for different parameter values of $\alpha$, highlighting that for \textit{a posteriori} credible values of $\alpha$ (gray shaded areas), the predictions are clearly distinct from the predictions of a WTA strategy (which assigns probability 1 to the highest scoring option for each item).
This suggests that the WTA strategy, which is the special limiting case for $\alpha \rightarrow \infty$, provides a substantially worse fit for item-level choices than those provided by less extreme values of $\alpha$.

@@

Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated LLM scores are inadequate predictors of the human data.
To corroborate this conclusion, Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM-based model predicts variance which is not borne out by the human data.
Concretely, the plots show, for each condition and item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
So, the model itself predicts systematic variability at the item level.
The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
Put more strongly, a model that uses the most obvious item-level score to predict what human participants choose on a by-item level is ruled out by the current experimental data.

@@
